### 2021-07-03 rad-icfp-2021/correctness.tex
bit more progress correctness

It is standard to give a denotational semantics of a first-order language in a Cartesian category.
An alternative approach consists in giving a semantics in a multicategory.
A multicategory generalises a category by allowing multimorphims, that is morphisms from a list of objects to an object.
Most categorical structures from category theory can be phrases similarly in multicategories.

@@

\begin{definition}{Syntactic multicategory for Source}
    Let $\SynSource$ be the multicategory whose objects are types of Source, and where a morphism 
    $[A_1,\ldots,A_n]\to A$ is a term $x_1:A_1,\ldots,x_n:A_n\vdash e:A$ of Source modulo the $\eta\beta$-laws.
    Composition is by substitution.
\end{definition}

@@

We can for instance use this to give a simple semantics of Source 
in the multicategory of Cartesian spaces and smooth maps between them.

\begin{definition}{CartSp}
    TODO, its definition and semantics of Source in it.
\end{definition}

@@

One main reason for introducing UNF is to have a better control 
of the computation flow of the term, similarly to what ANF or CPS do. 
A convenient categorical setting for this is to use string diagrams.
To better fit the standard denotational semantics of languages, 
we can instead use concategories.

Concategories are a generalisation of categories and multicategories in that 
they allow morphisms both from and to a list of objects. 
Similarly to categories and multicategories, 
most categorical abstractions used for the semantics of languages are easily adapted to them.

@@

\begin{example}{Concat1}
    TODO
\end{example}

We can interpret Source UNF in $\ConcatS$ as follows.

TODO

Similarly, we introduce a second concategory for the Target part.

\begin{example}{Concat2}
    TODO
\end{example}

@@

TODO

@@

We interpret Source in a new multicategory, 
whose morphisms are particular morphisms of $\ConcatS$.
As Source UNF is itself interpreted in $\ConcatS$, 
this gives us a way to compare terms in Source with terms in Source UNF.
This comparison precisely give the UNF transformation.

\begin{definition}{Mutlicat from concat}
    TODO
\end{definition}

\begin{proposition}{construction above gives UNF}
    TODO
\end{proposition}

@@

\begin{definition}{Concat from multicat}
    TODO
\end{definition}

\begin{proposition}{Initiality of concat2 gives $UNF^-1$}
    TODO
\end{proposition}

What remains to explain now is the reverse mode transformation between Source UNF and Target UNF.
We construct a functor $D:\ConcatS\to\ConcatT$. 
This functor will easily be shown to compute reverse mode derivatives.
Because terms of Source UNF in $\ConcatS$, we observe the effect of $D$ on them
and show that it matches the syntactic $\Dsynrevsymbol_{\rho}$ 
from Section~\ref{sub:Simple reverse mode transformation}.

\begin{definition}{D as a lax functor}
    TODO
\end{definition}

\begin{proposition}{semantic of syntactic D matches D lax functor}
    TODO
\end{proposition}

@@

First, we start from the correctness of $D$ which is easy to establish, 
and then somehow propagate this information to Source and Target.

\begin{proposition}{Correctness $D$}
    TODO
\end{proposition}

Recall that the intuition from Source UNF is that it consist of terms of Source that also return their context.
From there, the intuition for the Jacobian of a primitive in Target UNF is that it should be the Jacobian of
the corresponding term in Target.

@@

\begin{proposition}{The transpose Jacobian of !map2! and !reduce! are correct.}
    TODO
\end{proposition}

From this, we can now deduce that the composite transformation doing UNF then $D$ and then $UNF^-1$ is correct
in the sense that it produces a term which computes the gradient of the original term.

\begin{theorem}{UNF;D;$UNF^-1$ correct}
    TODO
\end{theorem}

By inspection of what that composition actually give on terms from Source, 
we show that the previous composition of transformation indeed gives the transformation 
$\directD{\rho}{\Gamma}{Y}$ from Section~\ref{sub:Macro for pure reverse mode transformation}.

\begin{proposition}{UNF;D;$UNF^-1$=D}
    TODO
\end{proposition}

@@

\begin{theorem}{D is correct}
    TODO
\end{theorem}

>>>>

It is standard to give a denotational semantics of a first-order language in a Cartesian category.
An alternative approach consists of giving semantics in a multicategory.
A multicategory generalizes a category by allowing multimorphisms, that is, morphisms from a list of objects to an object.
Most categorical structures from category theory can be phrases similarly in multicategories.

@@

\begin{definition}[Syntactic multicategory for Source]
    Let $\SynSource$ be the multicategory whose objects are types of Source, and where a morphism 
    $[A_1,\ldots,A_n]\to A$ is a term $x_1:A_1,\ldots,x_n:A_n\vdash e:A$ of Source modulo the $\eta\beta$-laws.
    Composition is by substitution.
\end{definition}

@@

We can, for instance, use this to give a simple semantics of Source 
in the multicategory of Cartesian spaces and smooth maps between them.

\begin{definition}[CartSp]
    TODO, its definition and semantics of Source in it.
\end{definition}

@@

One main reason for introducing UNF is to have a better handle over
the computation flow of the term, similar to what ANF or CPS provide. 
A convenient categorical setting for this is to use string diagrams.
To better fit the standard denotational semantics of languages, 
we can instead use concategories.

Concategories are a generalization of categories and multicategories in that 
they allow morphisms both from and to a list of objects. 
Similarly to categories and multicategories, 
most categorical abstractions used for the semantics of languages are easily adapted to them.

@@

\begin{definition}[$\ConcatS$]
Let $\ConcatS$ the syntactic Cartesian concategory whose types are those of Source and 
with primitives given by $op1: \reals \to \reals$, $op2: \reals,\reals \to \reals$ $map2_{x,y.e}$
and $reduce_{x,y.e1;e2}$.
\MH{actually I just need projection and a representing object for pairs, it might be weaker than Cartesian?}
\end{definition}

One may notice that the syntax of $\ConcatS$ is somewhere in between the syntax of Source and of Source UNF.
We can interpret Source UNF in $\ConcatS$ as follows.

\begin{tabular}{l c r}
   $\seml$!var!$_{T;i} \semr$ &=& \\
   $\seml$!op!$_{T;n} \semr$ &=& \\
   $\seml$!pair!$_{T;A\times B} \semr$ &=& \\
   $\seml$!proj!$_{T1;T2;T3} \semr$ &=& \\
   $\seml$!e!$\comp$!e!$\semr$  &=& \\
   $\seml$!e!$\pcomp$!e!$\semr$ &=& \\
   $\seml$!map2!$_{T;x,y.e}\semr$  &=& \\
   $\seml$!reduce!$_{T;x,y.e;e}\semr$  &=& \\
\end{tabular}

Similarly, we introduce a second concategory for the Target part of our transformations.

\begin{definition}[$\ConcatT$]
Let $\ConcatT$ be the syntactic Cartesian concategory whose types are those of Target and 
with primitives given by $op1: \reals \to \reals$, $op2: \reals,\reals \to \reals$ $map2_{x,y.e}$, 
$reduce_{x,y.e1;e2}$, TODO
\MH{not too sure, should I say what a Cartesian concat is?}
\end{definition}

@@

\begin{tabular}{l c r}
    $\seml$!var!$_{T;i} \semr$ &=& \\
    $\seml$!op!$_{T;n} \semr$ &=& \\
    $\seml$!pair!$_{T;A\times B} \semr$ &=& \\
    $\seml$!proj!$_{T1;T2;T3} \semr$ &=& \\
    $\seml$!e!$\comp$!e!$\semr$  &=& \\
    $\seml$!e!$\pcomp$!e!$\semr$ &=& \\
    $\seml$!map2!$_{T;x,y.e}\semr$  &=& \\
    $\seml$!reduce!$_{T;x,y.e;e}\semr$  &=& \\
 \end{tabular}
\MH{Do I need any kind of proposition saying that this makes sense?
Might need to define on types as well to be clearer: It's an identity on type functor...}

@@

We interpret Source in a new multicategory, 
whose morphisms are particular morphisms of $\ConcatS$.
As Source UNF is itself interpreted in $\ConcatS$, 
this gives us a way to compare terms in Source with terms in Source UNF.
This comparison precisely gives the UNF transformation.

\begin{definition}[Mutlicat from concat]
    TODO
\end{definition}

\begin{proposition}[construction above gives UNF]
    TODO
\end{proposition}

@@

\begin{definition}[Concat from multicat]
    TODO
\end{definition}

\begin{proposition}[Initiality of concat2 gives $UNF^-1$]
    TODO
\end{proposition}

What remains to explain now is the reverse mode transformation between Source UNF and Target UNF.
We construct a functor $D:\ConcatS\to\ConcatT$. 
This functor will easily be shown to compute reverse-mode derivatives.
Because terms of Source UNF in $\ConcatS$, we observe the effect of $D$ on them
and show that it matches the syntactic $\Dsynrevsymbol_{\rho}$ 
from Section~\ref{sub:Simple reverse mode transformation}.

\begin{definition}[D as a lax functor]
    TODO
\end{definition}

\begin{proposition}[semantic of syntactic D matches D lax functor]
    TODO
\end{proposition}

@@

First, we start from the correctness of $D$, which is easy to establish, 
and then propagate this information to Source and Target.

\begin{proposition}[Correctness $D$]
    TODO
\end{proposition}

Recall that the intuition from Source UNF is that it consists of terms of Source that also return their context.
From there, the intuition for the Jacobian of a primitive in Target UNF is that it should be the Jacobian of
the corresponding term in Target.

@@

\begin{proposition}[The transpose Jacobian of !map2! and !reduce! are correct.]
    TODO
\end{proposition}

From this, we can now deduce that the composite transformation doing UNF then $D$ and then $UNF^-1$ is correct
in the sense that it produces a term that computes the gradient of the original term.

\begin{theorem}[UNF;D;$UNF^-1$ correct]
    TODO
\end{theorem}

By inspection of what that composition actually gives on terms of Source, 
we show that the previous composition of transformation indeed gives the transformation 
$\directD{\rho}{\Gamma}{Y}$ from Section~\ref{sub:Macro for pure reverse mode transformation}.

\begin{proposition}[UNF;D;$UNF^-1$=D]
    TODO
\end{proposition}

@@

\begin{theorem}[D is correct]
    TODO
\end{theorem}

### 2021-07-03 rad-icfp-2021/relwork.tex
bit more progress correctness

Given the enormous computation needs for state-of-the-art large scale machine learning applications, 
which require extremely efficient tensor computations and automatic differentiation for backpropagation, 
combining array languages and automatic differentiation (tensor calculus) in the best 
fitted intermediate representation for optimisations is of key interest, and active research.
\cite{bernstein2020differentiating} 
\cite{laue2018computing,laue2020simple}

@@

\cite{ee2020correctness}: non diff function
\cite{sherman2021}: non diff HO functions
\cite{pearlmutter2008reverse}: based on this
\cite{elliott2018simple}: bit similar because using CT as useful language, but we take impl and opti more seriously.
\cite{shaikhha2019efficient}: similar approach but here more complicated for reversemode, would be interesting to investigate more loop fusions here.

@@

\clearpage

>>>>

Given the enormous computation needs for state-of-the-art large scale machine learning applications, 
which require extremely efficient tensor computations and automatic differentiation for backpropagation, 
combining array languages and automatic differentiation (tensor calculus) in the best 
fitted intermediate representation for optimizations is of key interest and active research.
\cite{bernstein2020differentiating} 
\cite{laue2018computing,laue2020simple}

@@

\cite{ee2020correctness}: non diff function
\cite{sherman2021}: non diff HO functions
\cite{pearlmutter2008reverse}: based on this
\cite{elliott2018simple}: bit similar because using CT as useful language, but we take impl and opti more seriously.
\cite{shaikhha2019efficient}: similar approach but here more complicated for reverse mode, would be interesting to investigate more loop fusions here.

@@

\MH{should talk a bit about very recent reverse AD in Dex \cite{paszke2021getting}}
\clearpage
